Topological Equivariant Artist Model

### 2021-03-31 paper/paper.tex
cleaned up intro

Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built \cite{wongsuphasawatNavigatingWideWorld2021}. These building block libraries provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools; because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle. To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software needs to satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be a structure preserving map from data to visual representation.

\begin{figure}
  \includegraphics[width=\columnwidth]{figures/continuity.png}
  \caption{The line plot does not preserve continuity because it implies that the discrete records are connected to each other, while the bar plot is continuity preserving because it visually represents the records as independent data points.}
  \label{fig:intro:continuity}
\end{figure}

The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}. As illustrated in \autoref{fig:intro:continuity}, \textit{continuity} is the way in which the records in the dataset are connected to each other. While the visualization tool can generate the line plot, it does not preserve continuity structure because it implies that the discrete categories are somehow continuous. 
\begin{figure}
  \includegraphics[width=\columnwidth]{figures/equivariance.png}
  \caption{The data in blue is scaled by a factor of two, yielding the data in orange. To preserve equivariance, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
  \label{fig:intro:equivariance}
\end{figure}
Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

Motivated by the challenge of rearchitecturing a general purpose visualization library, we developed the Topological Equivariant Artist Model (TEAM) to express the constraints of preserving data continuity and translating data properties into a visual representation. The contribution of this work is a  
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}

>>>>

Building block libraries underpin the visualization ecosystem; they are not the automated tools or high level libraries many practitioners use to build their visualizations, rather they are the libraries on which these tools are built. These building block libraries, as described by Wongsuphasawat, provide specific components and utilities that are composed by domain specific visualization practitioners into reusable visualization tools \cite{wongsuphasawatNavigatingWideWorld2021}. Because building block libraries often support many domains, the API can grow organically in ways that are incoherent and brittle. Motivated by the challenge of rearchitecturing a building blocks library, Matplotlib\cite{hunterMatplotlib2DGraphics2007}, we developed a functional model, which we call the Topological Equivariant Artist Model (TEAM), to formalize the constraints building block components must satisfy to be considered structure preserving maps from data to visual representation.

\begin{figure}[H]
  \centering 
  \subfloat[\label{fig:intro:continuity}]{%
    \includegraphics[width=1\columnwidth]{continuity.png}}\\
  \subfloat[\label{fig:intro:equivariance}]{%
    \includegraphics[width=1\columnwidth]{equivariance.png}}
\caption{In \autoref{fig:intro:continuity}, the line plot does not preserve \textit{continuity} because it implies that the discrete records are connected to each other, while the bar plot is \textit{continuity} preserving because it visually represents the records as independent data points. In \autoref{fig:intro:equivariance}, the data in blue is scaled by a factor of two, yielding the data in orange. To preserve \textit{equivariance}, the blue line plot representation of the unscaled data is also scaled by a factor of two, yielding the orange line plot that is equivalent to the scaled data.}
\label{fig:intro:structure}
\end{figure}
The structure that visualizations must preserve is \textit{continuity} and \textit{equivariance}, as illustrated in \autoref{fig:intro:structure}. \textit{continuity} is the way in which the records in the dataset are connected to each other. In \autoref{fig:intro:continuity} the line plot does not preserve continuity because the lines connecting the discrete categories implies that weather events and frequency of those events is continuous. Equivariance means that if any action is applied to the data or the visual, for example a rotation, permutation, translation, or rescaling, an equivalent action is applied on the other side. In \autoref{fig:intro:equivariance}, the data and visual representation are scaled by equivalent factors of two, resulting in the change illustrated in the shift from orange to blue.

To adapt to modern domain library development needs, this paper presents a functional model of the constraints building block software must satisfy so that a visualization--however it is specified in the library built out of these blocks--is guaranteed to be an equivariant continuity preserving mapping from data to visualization. The contribution of this work is 
\begin{enumerate}
  \item formalization of the topology preserving relationship between data and graphic via continuous maps \autoref{sec:math:graphic:base}
  \item formalization of property preservation from data component to visual representation as monoid action equivariant maps \autoref{sec:math:artist}
  \item functional oriented visualization architecture built on the mathematical model to demonstrate the utility of the model \autoref{sec:math:artist:q}
  \item prototype of the architecture built on Matplotlib's infrastructure to demonstrate the feasibility of the model. \autoref{sec:code}
\end{enumerate}
